Constructing Sailing Match Race Schedules: Round Robin Pairing Lists

### 2015-03-24 paper.tex
Layout
Backslash after i.e.

\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item When there is an even number of flights, each skipper has the same number
    of port and starboard assignments.
\item When there is an odd number of flights, the first half of the skippers
    will have one more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of
    the next flight.
\item No skipper should have more than two consecutive port or starboard
    assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as
    equally as possible.
\item In flights with five or more matches no skipper should be in the
    next-to-last match in a flight and then in the first match of the next
    flight.
\item If possible, a skipper should be starboard when meeting the nearest
    lowest ranked skipper (i.e. \#1 will be starboard against \#2, \#3 will be
    starboard against \#4).
\item Close ranked skippers meet in the last flight.
\item Minimize the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}

@@

The first thing we model is a skipper. A skipper $\sigma$ has a temporal view
of his schedule in the array of constrained integer variables $\timeSlot[0]$
to $\timeSlot[{t-1}]$ (equation \ref{eq3} below), and a state view in the array of
constraint integer variables $\state[0]$ to $\state[{f-1}]$ (equation \ref{eq4}
below), i.e. $\state[i]$ is the state of this skipper in flight $i$ and
corresponds to timeslots $\timeSlot[{i*m}]$ to $\timeSlot[{m(i+1)-1}]$.

@@

We must then ensure that each skipper $\sigma$ is in a match with all other
skippers $\{0, \ldots, n-1\} \setminus \{sigma\}$. This may be done by imposing
a global cardinality constraint \cite{globCard} on the array $\timeSlot$. The
$\state$ variables are then used to impose match race criteria 4 (if last in
flight $i$ then not first in flight $i+1$), criteria 11 (if last in flight $i$
then not in a bye in flight $i+1$) and criteria 12 (if flight $i$ is a bye then
flight $i$ is not first). In addition, we have the rule that if a skipper has
finished his matches in flight $i$ (i.e. END) then he has finished his matches
in flight $i+1$ (and so on).  These criteria are imposed on the $\state$
variables via a deterministic finite automata (DFA) constraint \cite{Pesant04}.

@@

We are now in a position to model criteria 1 (each skipper sails against every
other skipper) and optimisation criteria 10 (minimise boat changes).  In
\ref{eq13}, \ref{eq14} and \ref{eq15} we present a match perspective of the
schedule, using a two dimensional array of variables $\match$, where
$\match[{i,j}]$ is the time slot that skippers $s[{i}]$ and $s[{j}]$ meet in a
match. Only the half above the diagonal is represented, and the lower half is
made up of exactly the same variables, i.e. $\match[{i,j}]$ is \emph{exactly}
the same constrained variable as $\match[{j,i}]$. Constraint \ref{eq15} states
that  a match between skippers $s[{i}]$ and $s[{j}]$ takes place at time $k$
(i.e. $\match[{i,j}] = k$) if and only if  skipper $s[{i}]$'s $k^{th}$ time
slot is skipper $s[{j}]$ and conversely that skipper $s[{j}]$'s $k^{th}$ time
slot is skipper $s[{i}]$).  Constraints \ref{eq16}, \ref{eq17} and \ref{eq18}
then convert this from time slots to flights, i.e.  $\modMatch[{i,j}]$ is the
flight in which skippers $s[{i}]$ and $s[{j}]$ meet for a match. Finally,
constraint \ref{eq19} ensures that each skipper's match occurs in different
flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows
to be all different also ensures that columns are all different.

@@

We now take a temporal view (\ref{eq20}), such that $\timeVar[{\tau}]$ is a pair
$(i,j)$, stating that at time $\tau$ skippers $s[{i}]$ and $s[{j}]$ are in a
match, i.e. we channel between the match-perspective and the
temporal-perspective.
%
% one dimensional time and decision variables
%
\begin{flalign}
    & \forall_{\tau \in [0..t-1]} ~ \timeVar[{\tau}] \in \{(0,1) .. (n-2,n-1)\} \label{eq20} \\
    & \forall_{\tau \in [0 .. t-1]} ~ \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~ \timeVar[{\tau}] = (i,j) \iff \match[{i,j}] = \tau \label{eq21}
\end{flalign}

@@

\begin{figure}[h]
\centering
\includegraphics[height=10.2cm,width=13.2cm]{schedule.pdf}
%\vspace{-35mm}
\caption{A pictorial representation of the entire model of the schedule for 7 skippers and 6 boats. The schedule is reproduced top right. On the right we have the 7 skippers, on top left the $\modMatch$ array, at the bottom $\timeVar$ and middle left the $\match$ array. The yellow arrows signify channeling between parts of the model.}
\label{schedChanges} 
\end{figure}

The decision variables are $\timeVar[{0}]$ to $\timeVar[{t-1}]$, i.e. for each time slot we decide what match takes place. A symmetry break is made at the top of search by forcing the first round to contain the matches (0,1), (2,3), ... i.e. $\forall_{i \in [0..m-1]} ~ \match[{2.i,2i+1}] = i$. The schedule presented has 6 boat changes: there are no boat changes in flight 0 (first flight), flight 1 skipper 6 has a boat change, flight 2 skipper 3 has a boat change, flight 3 it's skipper 4, flight 4 skipper 6, flight 5 skipper 0 and last flight skipper 1.

\subsection{Stage 2 and 3: minimising imbalance and renaming skippers}
Having produced a schedule that minimises boat changes, we then minimise criteria 6, imbalance. Assuming the first stage produced a schedule with $\beta$ boat changes we post this as a hard constraint in stage 2, i.e. $\totalBoatChanges \leq \beta$. For each skipper $\sigma$ we introduce a zero/one variable $position[{i,j}]$  (\ref{posA}) which is 1 if and only if skipper $\sigma$ is in a match in time $m.j + i$, i.e. in the $i^{th}$ match of the  $j^{th}$ flight (\ref{posB}). Imbalance is then computed via (\ref{imbA}) for each position, as the absolute amount of over or under expected presence in a given postion. Finally, (\ref{imbB}) we compute the maximum imbalance over all positions for this skipper.

>>>>

\begin{enumerate}
\item Each skipper sails against each other skipper once.
\item When there is an even number of flights, each skipper has the same number
    of port and starboard assignments.
\item When there is an odd number of flights, the first half of the skippers
    will have one more starboard assignment.
\item No skipper in the last match of a flight should be in the first match of
    the next flight.
\item No skipper should have more than two consecutive port or starboard
    assignments.
\item Each skipper should be assigned to match 1, match 2, etc. in a flight as
    equally as possible.
\item In flights with five or more matches no skipper should be in the
    next-to-last match in a flight and then in the first match of the next
    flight.
\item If possible, a skipper should be starboard when meeting the nearest
    lowest ranked skipper (i.e.\ \#1 will be starboard against \#2, \#3 will be
    starboard against \#4).
\item Close ranked skippers meet in the last flight.
\item Minimize the number of boat changes.
\item Skippers in the last match of a flight do not change boats.
\item Skippers in new boats do not sail in the first match of the next flight.
\item Skippers have a reasonable sequence of matches and blanks.
\end{enumerate}

@@

The first thing we model is a skipper. A skipper $\sigma$ has a temporal view
of his schedule in the array of constrained integer variables $\timeSlot[0]$
to $\timeSlot[{t-1}]$ (equation \ref{eq3} below), and a state view in the array of
constraint integer variables $\state[0]$ to $\state[{f-1}]$ (equation \ref{eq4}
below), i.e.\ $\state[i]$ is the state of this skipper in flight $i$ and
corresponds to timeslots $\timeSlot[{i*m}]$ to $\timeSlot[{m(i+1)-1}]$.

@@

We must then ensure that each skipper $\sigma$ is in a match with all other
skippers $\{0, \ldots, n-1\} \setminus \{sigma\}$. This may be done by imposing
a global cardinality constraint \cite{globCard} on the array $\timeSlot$. The
$\state$ variables are then used to impose match race criteria 4 (if last in
flight $i$ then not first in flight $i+1$), criteria 11 (if last in flight $i$
then not in a bye in flight $i+1$) and criteria 12 (if flight $i$ is a bye then
flight $i$ is not first). In addition, we have the rule that if a skipper has
finished his matches in flight $i$ (i.e.\ $\END$) then he has finished his matches
in flight $i+1$ (and so on).  These criteria are imposed on the $\state$
variables via a deterministic finite automata (DFA) constraint \cite{Pesant04}.

@@

We are now in a position to model criteria 1 (each skipper sails against every
other skipper) and optimisation criteria 10 (minimise boat changes).  In
\ref{eq13}, \ref{eq14} and \ref{eq15} we present a match perspective of the
schedule, using a two dimensional array of variables $\match$, where
$\match[{i,j}]$ is the time slot that skippers $s[{i}]$ and $s[{j}]$ meet in a
match. Only the half above the diagonal is represented, and the lower half is
made up of exactly the same variables, i.e.\ $\match[{i,j}]$ is \emph{exactly}
the same constrained variable as $\match[{j,i}]$. Constraint \ref{eq15} states
that  a match between skippers $s[{i}]$ and $s[{j}]$ takes place at time $k$
(i.e.\ $\match[{i,j}] = k$) if and only if  skipper $s[{i}]$'s $k^{th}$ time
slot is skipper $s[{j}]$ and conversely that skipper $s[{j}]$'s $k^{th}$ time
slot is skipper $s[{i}]$).  Constraints \ref{eq16}, \ref{eq17} and \ref{eq18}
then convert this from time slots to flights, i.e.\ $\modMatch[{i,j}]$ is the
flight in which skippers $s[{i}]$ and $s[{j}]$ meet for a match. Finally,
constraint \ref{eq19} ensures that each skipper's match occurs in different
flights. Also, since $\modMatch[{i,j}] \equiv \modMatch[{j,i}]$, setting rows
to be all different also ensures that columns are all different.

@@

We now take a temporal view (\ref{eq20}), such that $\timeVar[{\tau}]$ is a pair
$(i,j)$, stating that at time $\tau$ skippers $s[{i}]$ and $s[{j}]$ are in a
match, i.e.\ we channel between the match-perspective and the
temporal-perspective.
%
% one dimensional time and decision variables
%
\begin{flalign}
    & \forall_{\tau \in [0..t-1]} ~ \timeVar[{\tau}] \in \{(0,1) .. (n-2,n-1)\} \label{eq20} \\
    & \forall_{\tau \in [0 .. t-1]} ~ \forall_{i \in [0..n-2]} ~ \forall_{j \in [i+1..n-1]}  ~ \timeVar[{\tau}] = (i,j) \iff \match[{i,j}] = \tau \label{eq21}
\end{flalign}

@@

\begin{figure}[tb]
\centering
\includegraphics[height=10.2cm,width=13.2cm]{schedule.pdf}
\caption{A pictorial representation of the entire model of the schedule for 7
skippers and 6 boats. The schedule is reproduced top right. On the right we
have the 7 skippers, on top left the $\modMatch$ array, at the bottom
$\timeVar$ and middle left the $\match$ array. The yellow arrows signify
channeling between parts of the model.}
\label{schedChanges}
\end{figure}

The decision variables are $\timeVar[{0}]$ to $\timeVar[{t-1}]$, i.e.\ for each
time slot we decide what match takes place. A symmetry break is made at the top
of search by forcing the first round to contain the matches $(0,1), (2,3), \ldots$,
i.e.\ $\forall_{i \in [0 \ldots m-1]} ~ \match[{2i,2i+1}] = i$. The schedule
presented has 6 boat changes: there are no boat changes in flight 0 (first
flight), flight 1 skipper 6 has a boat change, flight 2 skipper 3 has a boat
change, flight 3 it's skipper 4, flight 4 skipper 6, flight 5 skipper 0 and
last flight skipper 1.

\subsection{Stage 2 and 3: minimising imbalance and renaming skippers}
Having produced a schedule that minimises boat changes, we then minimise criteria 6, imbalance. Assuming the first stage produced a schedule with $\beta$ boat changes we post this as a hard constraint in stage 2, i.e.\ $\totalBoatChanges \leq \beta$. For each skipper $\sigma$ we introduce a zero/one variable $position[{i,j}]$  (\ref{posA}) which is 1 if and only if skipper $\sigma$ is in a match in time $m.j + i$, i.e.\ in the $i^{th}$ match of the  $j^{th}$ flight (\ref{posB}). Imbalance is then computed via (\ref{imbA}) for each position, as the absolute amount of over or under expected presence in a given postion. Finally, (\ref{imbB}) we compute the maximum imbalance over all positions for this skipper.
